\hypertarget{lambda-cdm}{%
\chapter{Lambda-CDM}\label{lambda-cdm}}

\hypertarget{evidence-for-dark-matter}{%
\section{Evidence for dark matter}\label{evidence-for-dark-matter}}

In 1932, Jan Oort studied the velocities of stars near the Sun, and he found
that the velocities of these stars were systematically larger than
expected~\cite{oort_force_1932}. From the velocities of stars at a given
radius in a galaxy, one can measure the gravitational mass of all objects
inside that radius. As such, Oort's velocity findings were really a statement
about the mass of the Milky Way. In Oort's own words, these velocities implied
that the amount of gravitating matter in the Milky Way must be larger than
what one can estimate by simply counting stars, with the latter falling short
by roughly 30-50\%~\cite{trimble_existence_1987}.

The next year, Fritz Zwicky studied velocity dispersions in the Coma galaxy
cluster~\cite{zwicky_rotverschiebung_1933,andernach_english_2017}. From his
measurements, he found that the velocity dispersions required ``10 to 100
times more mass'' for the cluster to remain bound than could be accounted for
by luminous matter. He concluded that there must be some large source of
non-luminous matter present to account for the difference, calling the
non-luminous matter ``dunkle Materie'', or ``dark matter''. This is often
cited as the origin of the term.

Over the next few decades, evidence continued to mount for the existence of
dark matter. Rotation curves began being considered, first by Babcock in
1939~\cite{babcock_rotation_1939} and Oort in 1940~\cite{oort_problems_1940}.
In the words of Oort, ``the distribution of mass in this object [the M31
galaxy] appears to bear almost no resemblance to that of light.'' Studies of
globular clusters in the Milky Way indicated that at least two-to-three times
as much mass must lie outside the orbit of the Sun as
inside~\cite{lohmann_masse_1956} (todo add Kunth 1952). Studies of binary
galaxies revealed large mass-to-light
ratios~\cite{page_average_1960,holmberg_masses_1954} (todo add van den Bergh
1961). More galaxy clusters were studied, and the trends noted by Zwicky were
found to hold generally~\cite{trimble_history_2013}. This mounting evidence largely
took the form of acknowledgement of a large mass-to-light ratio; it was not
widely thought that a yet unknown form of matter was responsible.

By the 1970s, however, that would change. This was in large part due to
developments on a few different fronts. First, the 1960s and 70s saw new
methods for measuring the masses of galaxy clusters. Both X-ray measurements
and gravitational lensing first began being used in the 1960s to obtain
measurements of the masses of galaxy clusters. These studies independently
confirmed the large, non-luminous masses that were indicated by the previous
velocity analyses. todo add a citation

Second, the cosmic microwave background was discovered in 1965 (todo add a
citation), and found to be remarkably isotropic (to better than a part in
\(10^4\)). However, strictly baryonic models of the universe suggest
anisotropies in the CMB on the order of \(\sim 3 \times 10^{-4}\). If the
Universe contained some non-baryonic matter which interacted with photons only
gravitationally, however, the predicted anisotropies are revised downward,
closer to \(10^{-5}\), consistent with observation. Dark matter seems to fit
this role perfectly. Anisotropies on this scale have been observed by more
recent studies, beginning with the COBE satellite in 1992 (todo add citation).

Third, data from rotation curves became quite convincing. Rubin and Ford
published improved optical data from the galaxy M31, the same galaxy studied
by Oort in 1940~\cite{rubin_rotation_1970}. They found that, to even farther
radii than had been previously considered, the rotation curve did \emph{not}
decrease outside the optically bright region of the galaxy, an even stronger
indicator of the existence of mass not accounted for by luminous matter.
Similarly, Roberts and Rots~\cite{roberts_comparison_1973} and Roberts and
Whitehurst~\cite{roberts_rotation_1975} used measurements of 21cm hydrogen
emissions to draw very similar conclusions.

Lastly, 1974 saw two watershed papers by independent research groups, each
concluding that galaxies have dark halos. Einasto, Kaasik, and Saar wrote
``the mass of galactic coronae halos exceeds the mass of populations of known
stars by one order of magnitude, as do the effective
dimensions''~\cite{einasto_missing_1974}. Similarly, Ostriker, Peebles, and
Yahil wrote ``the very large mass-to-light ratio and the very great extent of
the spiral galaxies can perhaps most plausibly be understood as due to a giant
halo of faint stars''~\cite{ostriker_size_1974}. They also note that the year
prior, Ostriker and Peebles had come to a similar conclusion from
considerations of stability, finding the traditional disk galaxy model to be
unstable without a spherical disk~\cite{ostriker_numerical_1973}.

\hypertarget{lambda-cdm-1}{%
\section{Lambda-CDM}\label{lambda-cdm-1}}

Since these discoveries, it has been widely believed among astronomers and
cosmologists that non-baryonic dark matter exists, that it comprises large,
spherical halos around galaxies, and that it played an integral role in the
formation of the structures we can see today. Evidence has become increasingly
strong with improved measurements of the anisotropy in the CMB (todo cite
someone), improved rotation curves for more galaxies (todo cite someone), and
better X-ray and gravitational lensing measurements.

Another major question of the twentieth century which came to be largely
settled toward its end was that of the cosmological constant, \(\Lambda\).
This constant was introduced by Einstein into his theory of general relativity
(todo explain why). Physically, this constant can be interpreted as the vacuum
energy density of empty space. It is often called ``dark energy'' for this
reason~\cite{schneider_extragalactic_2015}. In the late 1990s, measurements of
Type Ia supernovae strongly constrained \(\Lambda\) to be positive, giving
empty space a significant, constant vacuum
energy~\cite{riess_observational_1998,perlmutter_measurements_1999}.

These findings, combined with large amounts of evidence for the Big Bang
hypothesis, are all most simply described by the \(\Lambda\)CDM model,
sometimes called the standard model of cosmology or the concordance model~\cite{dodelson_modern_2021}. The model describes the Universe as consisting
of three components---the cosmological constant \(\Lambda\), cold dark matter
(CDM), and standard matter---operating under general relativity and coming
from an origination event roughly 14 Gyr ago (the Big Bang). This rather
simple model is remarkably successful at explaining the Universe.

Through its inclusion of the Big Bang event, the \(\Lambda\)CDM model gives an
explanation for the origin of the CMB. In this model, the CMB is a residual
radiation left-over from a period shortly after the Big Bang, where the
Universe was hot (\textgreater{}10,000 K). As the Universe cooled and
structures began to form, the photons from this period began to propagate
freely in all directions.

The \(\Lambda\)CDM model also provides a well-tested account of the formation
of structures in the Universe~\cite{dodelson_modern_2021}. In the early
Universe, small gravitational perturbations caused matter to collapse into
small structures. These small structures---effectively just perturbations in
the mass density of the Universe---created gravitational potential wells,
attracting other small structures and assembling together to produce larger
ones. This works especially well for describing the distribution of dark
matter in the Universe, as small pockets of dark matter converge to form
subhalos, halos, and larger. The case for baryonic matter and galaxy formation
is more complicated, involving baryonic processes, gas dynamics, and more, but
is also well-explained in the régime of hierarchical structure formation.
Simulations of hierarchical structure formation have found that it is able to
account well for the observed distribution of structures on the scales of
galaxies, galaxy clusters, and larger. The implications of this model of
structure formation are considered in more detail in the next section.

In fact, the success of this theory of structure formation serves as reason to
believe that dark matter is cold (hence the ``C'' in \(\Lambda\)CDM). Dark
matter being ``cold'' means that it was non-relativistic at the time of
decoupling (the period in the early Universe when matter began to fall out of
thermal equilibrium)~\cite{mohanty_astroparticle_2020}. By contrast, ``hot'' dark matter
(HDM) would have been relativistic at the time of decoupling. Given these
higher velocities, the velocity dispersion of HDM would have been
non-negligible (differing from the negligible velocity dispersion of CDM). A
finite velocity dispersion, however, would have prevented the dark matter
particles from being bound to shallow gravitational potential wells, further
preventing the formation of small-scale
structures~\cite{schneider_extragalactic_2015}. Thus, HDM fails to form
structure.

On its own, \(\Lambda\)CDM is remarkably able to explain most of our
observations of the Universe. However, there are a few cosmological problems
which have arisen. One is known as the horizon
problem~\cite{bergstrom_cosmology_2008,schneider_extragalactic_2015}, which
points out the the CMB is, to a high degree, isotropic, indicating that the
photons making up this radiation are roughly in thermal equilibrium across the
sky. However, this means that regions of the sky which are too far separated
to be causally connected have remained in equilibrium. This is not explained
by \(\Lambda\)CDM.

Another problem is the flatness problem, which points out that there is
potentially a fine-tuning problem with the vanilla \(\Lambda\)CDM
model~\cite{bergstrom_cosmology_2008,schneider_extragalactic_2015}. The
Universe as we observe it is approximately flat today, meaning that its
curvature is approximately unity. If the curvature were not unity, its
deviation from flatness is expected to grow with time. As such, if the
Universe is not flat, it must have been \emph{very} close to unity at early
times, meaning that this parameter may need to be ``fine-tuned'' to achieve a
reasonable explanation. The problem is that our model should be able to
describe parameters like these \emph{without} fine-tuning.

Combining the \(\Lambda\)CDM model with the theory of cosmic inflation,
however, solves both of these problems~\cite{bergstrom_cosmology_2008}.
Inflation posits that the Big Bang was followed by a period of rapid
(exponential!) inflationary expansion. Such a phenomenon would solve the
horizon problem, as the whole observable Universe would have originated from a
much smaller, thermally-connected region before inflation. It also solves the
flatness problem, as inflation would have significantly stretched and
flattened any curvature in the Universe, providing a natural explanation for
the observed flatness today. In fact, this creates a sort of inverse
fine-tuning problem: in this theory, any curvature which is \emph{far} from
unity requires fine-tuning.

\hypertarget{structure-formation}{%
\section{Structure formation}\label{structure-formation}}

As stated above, the \(\Lambda\)CDM model predicts structures to form in
the Universe as a result of gravitational perturbations at early times.
As expansion slows, the pockets of matter created by these perturbations
host small gravitational potential wells, causing nearby matter to
collapse inward and the growth of structure to occur. With the majority
of matter in the Universe being dark matter, the first structures which
occur are pockets of dark matter halos. The halos observable today are
the result of the hierarchical combination of smaller halos.

The abundance of these halos can be described by extended
Press-Schechter (EPS) theory~\cite{bullock_small-scale_2017}. This
theory is grounded on rather simple assumptions: that one can use a
spherical collapse model and that one can extrapolate from linear
perturbation theory even into non-linear regimes. Despite these
assumptions, it has been shown to give predictions about the mass
spectrum of dark matter halos which are in accord with high-resolution
numerical simulations.

The halos themselves can be described as virialized objects with mass
\begin{equation}
M_{\text{vir}} = \frac{4\pi}{3} R_{\text{vir}}^3 \Delta_c \rho_c,
\end{equation}
where \(R_{\text{vir}}\) is the virial radius, \(\rho_c\) is the
critical density of the Universe, and \(\Delta_c\) is the over-density
parameter. In this work, we choose to set \(\Delta_c = 200\) and thus
mark the virial mass and radius as with a subscript 200 instead of
``vir''.

The \(\Lambda\)CDM model also provides predictions about the internal
structure of these dark matter halos. Dark matter-only N-body simulations were
performed as early as 1988~\cite{frenk_formation_1988}, with higher-resolution
simulations like~\cite{dubinski_structure_1991}
and~\cite{navarro_structure_1995} following shortly thereafter. The resulting
halos from these simulations were found to agree remarkably well with a
\emph{two-power} density profile~\cite{binney_galactic_2008} 
\begin{equation} \label{eq:twopower}
\rho(r) = \frac{\rho_0}{(r/a)^\alpha (1 + r/a)^{\beta-\alpha}}, 
\end{equation}
where \(\rho_0\) is a characteristic density for the halo and \(a\) is the
length scale. Special cases of this model include \((\alpha,\beta) = (1,4)\),
the Hernquist model~\cite{hernquist_analytical_1990}, and \((\alpha,\beta) =
(1,3)\), the NFW model~\cite{navarro_structure_1995,navarro_universal_1997}.
In general, the NFW model enjoys the most usage as one of the most simple and
accurate models of the mass density distribution of smaller halos.

todo Include some figures

todo Maybe include a more complete derivation of enclosed mass, NFW virial mass,
hernquist total mass, etc.

The required parameters above, \(\rho_0\) and \(a\), are typically
reformulated for the NFW distribution. In particular, we can take a
given virial mass \(M_{200}\), determine the corresponding virial radius
\(R_{200}\), and define the halo concentration \(c = R_{200} / a\). The
characteristic density can be found by integrating the density profile
up to \(R_{200}\) and setting it equal to \(M_{200}\). In this way, the
virial mass and concentration are enough to completely specify the NFW
distribution. The result is given by
\begin{equation}
\rho_0 = \frac{M_{200}}{4 \pi a^3 f(c)}, \quad 
f(c) = \log(1+c) - c/(1+c).
\end{equation}

As these halos are formed hierarchically from smaller halos, however, it
has been found that the dense centers of the subhalos are able to
survive the merging process. A direct result of this finding is that
dark matter halos today should be full of substructure with satellite
subhalos of varying sizes. In fact, simulations have shown that the
number of subhalos within a halo is approximately self-similar for host
halo mass~\cite{bullock_small-scale_2017}. While it is difficult to
determine a method for identifying and counting these subhalos, it does
yield a testable prediction of the small-scale structure of dark matter
in the \(\Lambda\)CDM paradigm.

todo Discuss galaxy formation with an eye for how this leads to the
missing satellites and too-big-to-fail problems

\hypertarget{small-scale-problems}{%
\section{Small-scale problems}\label{small-scale-problems}}

The theory of structure formation that follows from \(\Lambda\)CDM leads
us to a few important results. First, we expect the density
distributions of dark matter halos to approximately follow the NFW
model, specifically with \(r^{-1}\) dependence for small \(r\). Second,
we expect rich substructure in massive dark matter halos with
predictions for the number of subhalos of varying masses contained
within. Third, through galaxy formation and the method of abundance
matching, we obtain predictions about the number and masses of satellite
galaxies that should be hosted in galaxies like the Milky Way. These
three key deductions form the basis for the three classic
\textbf{small-scale problems} in the \(\Lambda\)CDM model.

The first problem we will discuss is known as the \textbf{core-cusp problem},
which is primarily related to the first deduction above. Dark matter-only
simulations of CDM halos show that the expected distribution of mass follows a
density distribution that is very dense and \emph{cuspy}, i.e.~one that goes
like \(r^{-1}\) at small radii (Bullock and Boylan-Kolchin 2017). As early as
1993, however, it was pointed out that such a distribution is inconsistent
with observational data. In particular, ``{[}cuspy{]} density profiles are
excluded by gravitational lensing analyses on cluster scales and by the
rotation curves of gas-rich, halo-dominated dwarf spirals on small
scales''~\cite{flores_rotation_1993}. It was also recognized by Navarro,
Frenk, and White, who wrote ``CDM halos are too concentrated to be consistent
with the halo parameters inferred for dwarf
irregulars''~\cite{navarro_structure_1995}.

The second problem is the \textbf{missing satellites} problem.  $\Lambda$CDM
simulations predict that galaxies as large as the Milky Way should have as
many as \(\sim 1~000\) dark subhalos large enough to host dwarf
galaxies~\cite{bullock_small-scale_2017}.  However, as of 2019, less than 60
dwarf galaxies are known within the Milky Way~\cite{simon_faintest_2019}.

There exists a potential solution to the missing satellites problem,
however, in the form of abundance matching. First, one can expect that
dark matter halos become less efficient at making galaxies with
decreasing mass, and therefore there exists some threshold mass below
which these halos remain completely dark. Then, abundance matching
allows one to ``solve'' the missing satellites problem for the subhalos
above this threshold~\cite{bullock_small-scale_2017}.

This solution to the missing satellites problem thus makes a testable
prediction. The central masses of Milky Way satellites should be
consistent with the central masses of the most massive subhalos in
\(\Lambda\)CDM simulation~\cite{bullock_small-scale_2017}. First
probed by Boylan-Kolchin et al.~in 2011 (todo cite), however, this does not
hold; the most massive subhalos predicted in simulation are significantly more
massive than the most massive satellite galaxies. If these subhalos do exist
but have just remained dark, then we have a new problem: why did these
subhalos fail to form galaxies? This is known as the \textbf{too-big-to-fail
problem}, as these subhalos ought to be too big to fail to form observable
galaxies. If these massive dark subhalos just do not exist, then this is a
fundamental failing of the \(\Lambda\)CDM model's predictions in line with the
missing satellites problem.

Some believe that these small-scale problems are the result of the omission of
baryonic effects in \(\Lambda\)CDM simulations. As described above, the
majority of the conflicting predictions come as the result of dark matter-only
simulations. Further, in recent years, it has been shown that all of the above
problems (and other small-scale issues) can be reduced or eliminated by the
inclusion of stellar feedback and other effects. For example,~\cite{pontzen_how_2012} show that supernova feedback can alleviate the core-cusp
problem, and~\cite{chan_impact_2015} show that stellar feedback can explain
the too-big-to-fail problem.

However, prior to these findings, researchers sought modifications to the
model that would preserve the successes of \(\Lambda\)CDM on large scales
\emph{and} solve the small-scale problems. Some researchers have sought
modifications to the model of gravity, but a more popular avenue has been the
modification of the dark matter model. While it seems that integration of more
sophisticated baryonic processes may be sufficient to solve the small-scale
problems, there is no evidence yet that other proposed dark matter models are
incorrect descriptions of the particle nature of dark matter. As such, further
tests of \(\Lambda\)CDM and these alternative dark matter models are necessary
as probes of dark matter particle physics. The alternative dark matter model
that we consider in this work is \emph{self-interacting dark matter}, or SIDM.
