\documentclass[12pt,oneside]{book}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[parfill]{parskip}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{caption,subcaption}
\usepackage{setspace}

\DeclareUnicodeCharacter{0394}{\ensuremath{\Delta}}
\DeclareUnicodeCharacter{039B}{\ensuremath{\Lambda}}
\DeclareUnicodeCharacter{03A9}{\ensuremath{\Omega}}
\DeclareUnicodeCharacter{2212}{\ensuremath{-}}

\usepackage[backend=biber,bibencoding=utf8,sorting=none,style=phys, maxbibnames=3]{biblatex}
\addbibresource{refsall.bib}

\usepackage{hyperref}

\title{Effects of dark matter self-interaction on the evolution of the Sagittarius stream}
\author{Connor Hainje\\\textit{Advised by} Professor Mariangela Lisanti}
\date{April 30, 2021}

\begin{document}

\maketitle

\onehalfspacing
\pagenumbering{roman}

\chapter*{Abstract}

I present a novel test of self-interacting dark matter (SIDM) models by
analyzing the impact of self-interaction on the infall of the Sagittarius
(Sgr) dwarf spheroidal galaxy.  A review of the history of dark matter
evidence and the development of the $\Lambda$CDM cosmological model is
included, followed by a review of the motivations for and viability of SIDM
models. Sgr is then presented as a testbed for these models, since its unique
orbital history is highly sensitive to the microphysics determining the Galactic
gravitational potential. I present three simulations of the Sgr infall: two
using collisionless dark matter with cuspy and cored initial galactic profiles,
and one using self-interacting dark matter.  It is found that the inclusion of
dark matter self-interaction has profound impacts on the resulting shape and
distribution of the Sgr stream.  In particular, we find that the distribution of
radial velocities is impacted substantially, with the SIDM merger unable to
reconcile both the shape of the stream and the observed radial velocity of the
progenitor. 

\vspace{0.25in}
This paper represents my own work in accordance with University regulations.

\vspace{0.1in}
\hspace{3.5in}/s/ Connor Hainje

\newpage

\newpage

\singlespacing
\tableofcontents 
\newpage

\pagenumbering{arabic}
\onehalfspacing

\include{tex/Lambda-CDM}
\include{tex/SIDM}
\include{tex/Sgr}
\include{tex/Setup}
\include{tex/Sims}
\include{tex/Conclusions}

\setcounter{secnumdepth}{-1}
\chapter{Acknowledgements}

First and foremost, I would like to thank my advisor, Professor Mariangela
Lisanti, for a fantastic and fulfilling year of research.  This research has
been incredibly exciting and interesting for me, and I always felt you left
plenty of room for me to shape things how I would like.  Without your guidance
and advice, however, none of this would have been possible, and for that I am
very grateful.

I would also like to thank the friends in my life who have helped support me
in the long journey it took to get to where I am today.  I would like to
specifically thank my home friends---Curtis, Nate, Trumond, and Sara---for
helping me to get \textit{to} Princeton, and my Princeton friends---Alvin,
Karolen, Simon, and Jason---for helping me to get \textit{through} Princeton.

Next, I want to thank Kathryn.  You have been an immense source of happiness
and support to me for the past several years, and I feel you have helped me to
grow a lot.  Specifically, though, I want to thank you for helping me push
through the rough times and to enjoy the happy ones on the other side.

Lastly, to my family: Mom, Dad, Justin, Brianna, and GJ\&GJ, thank you always
being there and supporting me as I took on new challenges and pushed myself to
new heights.  Thank you also for helping to mold me into the person I am
today.  I don't know what I would do without you all.

\chapter{Institutional acknowledgements}

This research relied heavily on NumPy\footnote{\url{https://numpy.org}}
\cite{harris_array_2020}, Matplotlib\footnote{\url{https://matplotlib.org}}
\cite{hunter_matplotlib_2007}, and Astropy\footnote{\url{https://www.astropy.org}}
\cite{astropy_collaboration_astropy_2013, astropy_collaboration_astropy_2018},
a collection of open-source Python packages for numerical computing, data
visualization, and astronomy.

The simulations presented in this article were performed on computational resources managed and supported by Princeton Research Computing, a consortium of groups including the Princeton Institute for Computational Science and Engineering (PICSciE) and the Office of Information Technology's High Performance Computing Center and Visualization Laboratory at Princeton University.

\printbibliography

\end{document}