\chapter{Simulations}
\label{sec:sim}

\section{Methodology}
The following is a description of our pipeline for simulating the infall of the
Sgr dSph into the Milky Way. 

We generate the initial distributions of stellar and dark matter using a package
called GalactICS~\cite{deg_galactics_2019}. The galaxy is modeled with a
stellar disk and a dark matter halo. The halo mass density follows a
Navarro-Frenk-White (NFW) distribution,~\cite{wang_equilibrium_2010} given by
\begin{equation} \label{eq:halo_dist}
    \rho_{\text{halo}}(r) 
    = \frac{M_{200}}{4\pi f(c) r_{200}} \frac{cx}{r^2 (1+x)^2},
\end{equation}
where $M_{200}$ and $r_{200}$ are the Virial mass and radius, $c$ is the
concentration parameter, $r$ is the spherical radius, $f(c) \equiv \ln(1+c) -
c/(1+c)$, and $x \equiv rc/r_{200}$.

The disk mass density follows the following distribution
from~\cite{wang_equilibrium_2010}
\begin{equation} \label{eq:disk_dist}
    \rho_{\text{disk}}(R, z)
    = \left( \frac{c_0^2 M_{\text{disk}}}{4\pi} \right)
    \frac{
        b_0 R^2 + (b_0 + 3 \sqrt{z^2 + c_0^2})(b_0 + \sqrt{z^2 + c_0^2})^2
    }{
        \left[ R^2 + (b_0 + \sqrt{z^2 + c_0^2})^2 \right]^{5/2}
        (z^2 + c_0^2)^{3/2}
    },
\end{equation}
where $R$ is the cylindrical radius in the plane of the disk, $z$ is the
distance from the disk's plane, $M_{\text{disk}}$ is the mass of the disk, $b_0$
is the disk scale-radius, and $c_0$ is the disk scale-height.

Both of these density distributions are subject to truncation beyond a given
$r_{t}$ with a width of $dr_t$, where the truncation function is given
by~\cite{widrow_dynamical_2008}
\begin{equation} \label{eq:trunc}
    C(r; r_t, dr_t)
    = \frac{1}{2}\ \text{erfc} \left( \frac{r - r_t}{\sqrt{2}
    dr_t} \right).
\end{equation}
The disk and the halo have different values for $r_t$ and $dr_t$, and each
density distribution can be multiplied by the truncation function with
corresponding truncation parameters to obtain the ``true'' density distribution.

The parameters used for the initial MW and Sgr galaxies are detailed in
Table~\ref{tab:params}, who have compiled these results from a review of
existing literature and simulations of their own.
The GalactICS initial conditions are then converted to a GADGET binary initial
conditions file for use in
GIZMO~\cite{hopkins_gizmo_2015,springel_cosmological_2005}---the conversion is
handled by a GalactICS subpackage. SIDM runs use a public GIZMO subpackage
from~\cite{robles_sidm_2017}.

\begin{table}
    \centering
    \begin{tabular}{llll}
        \hline\hline
        Parameter                       & & MW & Sgr dSph \\ 
        \hline
        Halo total mass                 & $M_{\text{halo}}$
                                        & $1.25 \times 10^{12}$ M$_\odot$ 
                                        & $1.3 \times 10^{10}$ M$_\odot$ \\
        Halo concentration parameter    & $c$
                                        & $10$ 
                                        & $8$ \\
        Halo Virial mass                & $M_{200}$
                                        & $1 \times 10^{12}$ M$_\odot$
                                        & $1 \times 10^{10}$ M$_\odot$ \\
        Halo Virial radius              & $r_{200}$
                                        & $206$ kpc 
                                        & $44$ kpc \\
        Halo scale-radius               & $r_D$
                                        & ?? 
                                        & ?? \\
        Number of halo particles        & $N_{\text{halo}}$
                                        & $1.16 \times 10^6$ 
                                        & $1.17 \times 10^4$ \\
        \hline
        Disk total mass                 & $M_{\text{disk}}$
                                        & $8.13 \times 10^{10}$ M$_\odot$ 
                                        & $7.8 \times 10^{10}$ M$_\odot$ \\
        Disk scale-length               & $b_0$
                                        & $3.5$ kpc 
                                        & $0.85$ kpc \\
        Disk scale-height               & $c_0$
                                        & $0.53$ kpc 
                                        & $0.13$ kpc \\
        Number of disk particles        & $N_{\text{disk}}$
                                        & $2.03 \times 10^6$
                                        & $1.95 \times 10^4$ \\ 
        \hline\hline
    \end{tabular}
    \caption{Initial conditions parameters for the Milky Way (MW) and
    Sagittarius (Sgr dSph) galaxies. Many of these values are taken from~\cite{dierickx_predicted_2017}.}
    \label{tab:params}
\end{table}

In order to ensure that the initial galaxies are in equilibrium, we first
evolve each galaxy for 10 Gyr using GIZMO, each at rest and centered at the
origin. This has uncovered some surprising trends, as the mass distribution
of the galaxies has a tendency to evolve away from the initial NFW and
exponential distributions. Further discussion of these trends will follow in
Chapters \ref{sec:cdm-sim} and \ref{sec:sidm-sim}.

After equilibration, the final snapshots for the Milky Way and Sagittarius
galaxies are merged. The particles from the MW are centered at the origin
with no initial velocity. The Sagittarius galaxy is centered at $\vec{r} =
(125, 0, 0)$ kpc, with an initial velocity of $\vec{v} = (-10, 0, 70)$
km~s$^{-1}$. The merged snapshot is then used as an initial conditions GIZMO
file and evolved for another 10 Gyr. Snapshots are taken approximately every
0.1 Gyr. In most of these simulations, conditions similar to today are found
after between 6 and 8 Gyrs of evolution.

\input{tex/s4-cdm-sim}
\input{tex/s5-sidm-sim}